---
title: "Multi-sample analysis of human melanoma data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Multi-sample analysis of human melanoma data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

This tutorial provides an example using POLYspace on human melanoma data from Hoch et al., 2022 to analyze and compare multiple spatial sequencing samples.

Before running this tutorial, please make sure that the POLYspace package is installed.


```{r setup}
library(POLYspace)
```

The human melanoma spatial sequencing data used in this tutorial are included in the package and can be loaded directly using:
```{r}
data(dataset_melanoma)
names(dataset_melanoma)
```
The storage contains the following elements:

- `samples`: A list containing the required input for each sample, including cellular features and spatial coordinates, which serves as the primary input for downstream analyses.

- `sample_id`: A unique identifier for each spatial sample, used to distinguish different slides in multi-sample analyses.

- `group`: Sample-level grouping information indicating immune infiltration status, used for comparative analyses across conditions.

- `subject_id`: The identifier of the patient from whom each sample was obtained, allowing multiple samples from the same individual to be linked.

- `des`: The experimental design matrix specifying the relationships among samples, groups, and subjects for regression analyses comparing cellular neighborhoods across multiple slides.

- `coef`: The name(s) of the model coefficient(s) defining the contrast(s) of interest in the multi-sample regression analyses.



```{r}
samples = dataset_cortex$samples
sample_id = dataset_cortex$sample_id
```

### Create a POLYspace object

Because this dataset contains a large number of samples, we recommend not combining all samples into a single POLYspace object. 
Instead, we suggest constructing one POLYspace object per sample, which allows enrichment analyses to be parallelized across samples. 
This strategy enables more efficient use of computational resources and substantially improves overall analysis performance.
Below we illustrate the analysis pipeline for a single sample, from input construction through enrichment analysis, following the same workflow used for the mouse cortex example.


```{r, eval = FALSE}
# Retrieve the task ID from the SLURM array environment variable
taskID = as.integer(Sys.getenv("SLURM_ARRAY_TASK_ID"))

# Create a POLYspace object for a single sample specified by the task ID
obj = createPOLYspaceObject(
  samples = samples[taskID],
  id_of_samples = sample_id[taskID]
)

# Construct the spatial cell–cell neighboring network
obj = constructSpatialNetwork(POLYspace = obj)

# Identify spatial neighborhoods based on the constructed network
obj = neighborhoodMining(POLYspace = obj)

# Annotate the neighborhoods
obj = neighborhoodAnnotation(POLYspace = obj)
```

After neighborhood annotation, we construct a consistent set of neighborhoods of interest across samples by taking the union of neighborhoods for each target.  
This ensures that the same set of neighborhoods is considered for all samples.  
The union is obtained using the following code, where \code{files} is a list of all annotated \code{POLYspace} objects.
```{r, eval = FALSE}
targets = files[[1]]@targets

neighborhoods_of_interest = lapply(targets, function(to){
  unlist(sapply(files,function(ob){
    ob@annotations[[1]][match(list(to),ob@targets)]
  }))

})
neighborhoods_of_interest = lapply(neighborhoods_of_interest,function(x) base::sort(unique(x)))
neighborhoods_of_interest = lapply(neighborhoods_of_interest,function(x) {
  x[!grepl("other", x)]
})

for(i in 1:length(targets)){
  if(!(identical(targets[[i]],'singlet')|identical(targets[[i]],'pair'))){
    vec = neighborhoods_of_interest[[i]]
    if(!is.null(vec)){
      jd_frame = lapply(vec, function(x){
      y = stringr::str_split(x,'--')[[1]]
      sapply(stringr::str_split(y,'_'), function(z) z[2])
      })
    
    
      jd = sapply(jd_frame, function(x) {
        
        length(unique(x)) == length(x)
        
      })
      
      neighborhoods_of_interest[[i]] = vec[jd]


    }
    

  }
}
```


```{r, eval = FALSE}
# Perform neighborhood enrichment analysis using permutation testing
obj = neighborhoodEnrichment(POLYspace = obj,
                             neighborhoods_of_interest = neighborhoods_of_interest,
                             permutation_times = 1000,
                             ncores = 50)
```

After enrichment analysis has been completed for all samples, the resulting \code{POLYspace} objects are collected into a list, denoted here as \code{files}.  
This list is then used as input to identify differential neighborhoods associated with immune infiltration status, while controlling for subject information.


```{r, eval = FALSE}
# Extract the design matrix for comparing immune-hot versus immune-cold samples
des = dataset_melanoma[["des"]][["hot_vs_cold"]]

# Extract the coefficient specifying the contrast of interest (hot vs. cold)
coef = dataset_melanoma[["coef"]][["hot_vs_cold"]]

# Perform joint enrichment analysis to identify differential neighborhoods
# associated with immune infiltration status while controlling for subject effects
res = polyEnrichment(
  object = files,
  neighborhoods_of_interest = "union",
  des = des,
  coef = coef
)


```

The parameter `neighborhoods_of_interest` specifies which set of neighborhoods is included in the joint analysis across samples. 
Because each sample may contain a different set of neighborhood topologies, this parameter controls how neighborhoods are aligned across samples.

- `"union"`: Uses the union of all neighborhood types observed across samples. 
  This option includes any neighborhood that appears in at least one sample and is useful when neighborhood composition varies substantially across samples.

- `"intersect"`: Uses only the neighborhoods that are shared by all samples. 
  This option restricts the analysis to common neighborhood types and may be preferable when focusing on highly consistent spatial patterns.



