---
title: "Visualize cell neighborhood glyphs"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Visualize cell neighborhood glyphs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
  
```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

This tutorial provides examples for visualizing cell neighborhood glyphs.
Before running this tutorial, please make sure that the POLYspace package is installed.


```{r setup}
library(POLYspace)
library(patchwork)

```

The following visualization takes neighborhood strings and their corresponding adjacency matrices as input. In this example, the glyphs are displayed as a list to illustrate different neighborhood topologies.

First, examples without region information are presented.
```{r}
string_list = c('CD8+ T cell','CD8+ T cell--Tumor','CD8+ T cell--pDC--Tumor')
adjmat_list = list('singlet',
                   'pair',
                   matrix(c(0,1,0,1,0,1,0,1,0),3,3))

```

Arc visualization of cell neighborhood glyphs without explicit region information:
```{r}
celltype_color_plate = c(
  "CD8+ T cell" = 'black',
  "Tumor"       = 'black',
  "pDC"         = 'black'
)
pl_list = lapply(1:length(string_list),function(x){
  
  pl = neighborhoodArcGlyph(string = string_list[x],
                             adjmat = adjmat_list[[x]],
                             celltype_color_plate = celltype_color_plate,
                             node_scale = list(c(0.4,0.2),
                                               c(0.2,0.2)),
                             xlims = c(-0.8,0.8),
                             ylims = c(-1,1))
  
})


combined = wrap_plots(pl_list, nrow = 1)
combined
```


Arc visualization with colors for cell types:
```{r}
celltype_color_plate = c(
  "CD8+ T cell" = "#F03C3C",
  "pDC"         = "#9F7FDB",
  "Tumor"       = "#5F6FD0"
)
pl_list = lapply(1:length(string_list),function(x){
  
  pl = neighborhoodArcGlyph(string = string_list[x],
                             adjmat = adjmat_list[[x]],
                             celltype_color_plate = celltype_color_plate,
                             node_scale = list(c(0.4,0.2),
                                               c(0.2,0.2)),
                             xlims = c(-0.8,0.8),
                             ylims = c(-1,1))
  
})


combined = wrap_plots(pl_list, nrow = 1)
combined
```


Sector visualization with colors for cell types:
```{r}

pl_list = lapply(1:length(string_list),function(x){
  
  pl = neighborhoodSectorGlyph(string = string_list[x],
                              region_color_plate = region_color_plate,
                              celltype_color_plate = celltype_color_plate,
                              adj_matrix = adjmat_list[[x]],
                              region_alpha = 0.4,
                              cell_size = 8,
                              text = FALSE,
                              text_size = 4)
  
})


combined = wrap_plots(pl_list, nrow = 1)
combined

```

Next, examples without region information are presented.
```{r}
string_list = c(
  "TME_CD8+ T cell",
  "TME_CD8+ T cell--Tumor_Tumor",
  "TME_CD8+ T cell--TME_pDC--Tumor_Tumor"
)

adjmat_list = list('singlet',
                   'pair',
                   matrix(c(0,1,0,1,0,1,0,1,0),3,3))
```

Arc visualization of cell neighborhood glyphs with region information:
```{r}
celltype_color_plate = c(
  "CD8+ T cell" = 'black',
  "Tumor"       = 'black',
  "pDC"         = 'black'
)

region_color_plate = c(
  "Tumor" = "#7295D1",   
  "TME"   =  "#B0DDD3" 
)
pl_list = lapply(1:length(string_list),function(x){
  
  pl = neighborhoodArcGlyph(string = string_list[x],
                           adjmat = adjmat_list[[x]],
                           celltype_color_plate = celltype_color_plate,
                           region_color_plate = region_color_plate,
                           node_text_size = 5, 
                           node_scale = list(c(0.4,0.2),
                                             c(0.2,0.2)),
                           xlims = c(-0.8,0.8),
                           ylims = c(0,4))
  
})


combined = wrap_plots(pl_list, nrow = 1)
combined
```


Sector visualization of colored cell types with region information
```{r}
celltype_color_plate = c(
  "CD8+ T cell" = "#F03C3C",
  "pDC"         = "#9F7FDB",
  "Tumor"       = "#5F6FD0"
)
pl_list = lapply(1:length(string_list),function(x){
  
  pl = neighborhoodSectorGlyph(string = string_list[x],
                              region_color_plate = region_color_plate,
                              celltype_color_plate = celltype_color_plate,
                              adj_matrix = adjmat_list[[x]],
                              region_alpha = 0.4,
                              cell_size = 8,
                              text = FALSE,
                              text_size = 4)
  
})


combined = wrap_plots(pl_list, nrow = 1)
combined

```
